Road Segment Prioritization for Bicycle Infrastructure

### 2021-02-26 README.Rmd
text edits

*Optimization* techniques have been used to propose improvements to
cycling networks. @mesbah2012bilevel propose a bi-level formulation to
optimize allocation of cycling lanes to the network without exceeding a
set budget. They account for the effect of cycling lanes on car traffic,
and attempt to maximize utilization of said lanes with minimal impact on
car travel times. Safeguarding against increased car traffic may be counter-productive if the goal is to create a mode-shift, as research has shown that reducing road space for cars leads to less cars on the road, a phenomenon referred to as "traffic evaporation" [@nello2020environmental].
Car usage is not considered by @mauttone2017bicycle,
who develop an optimization framework that aims to minimize the total
user cost of cycling on the network. The aggregate flow^[*flow* is
used throughout this research to refer to the cycling demand when it is
routed onto the road network. The flow on any road segment is the
cumulative demand on it, resulting from cyclists commuting between
various OD pairs] on links^[*link*
refers to a road segment throughout this research] is obtained by using
shortest paths to route existing cycling demand onto the road network,
and the solution is a proposed set of links where cycling infrastructure
should be added in order to minimize the overall travel cost of cyclists
across the network. The cost of traversing a link is given as a function
of its length and whether or not it has cycling infrastructure. The
problem has also been solved by attempting to find the minimum cost of
improving roadway links to meet a desired level of service (LOS)
[@duthie2014optimization]. In this formulation, all origin-destination
(OD) pairs need to be connected by roads that meet the desired LOS. A
limitation of these problem formulations do not explicitly solve for
continuity, which is dealt with using either a constraint specifying
that each link with a bike lane should be connected to at least one
destination [@mesbah2012bilevel], a constraint on maximum deviation
from shortest paths [@duthie2014optimization], or a discontinuity
penalty to prioritize connected road segments [@mauttone2017bicycle].

@@

All of these network-level methodologies are underpinned by ethical
principles, even though these principles are not explicitly acknowledged
by the authors. This is important since different ethical principles
constitute different problem formulations and targets. Broadly speaking,
transport appraisal can be based on either utilitarian or egalitarian
principles. The former seeks to maximize the overall benefit, while the
latter is concerned with a fair distribution of benefits
[@jafino2020transport]. @nahmias2017integrating criticize the
utilitarian approach that has been historically popular in the
evaluation of transport investments, explaining how the maximization of
overall benefit fails to account for the distribution of that benefit
among communities or individuals. @lucas2016method explain how
transport studies have traditionally looked at the bigger picture
without studying the distribution of investments on the different parts
of the study area, and go on to propose an egalitarian approach that
ensures the disaggregation of transport policy benefits across the study
area. @pereira2017distributive also emphasize the need for a more
egalitarian approach to transport planning. They highlight accessibility
as a cornerstone of distributive justice, and contend that policies
should aim to distribute investments in a way that minimizes spatial
variations in accessibility.

@@

```{r potdemhistograms, fig.cap="Distribution of Potential Cycling Demand", out.width="32%", fig.show='hold', echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/histogram_distance_all_vs_cycling.png",
  "data/Manchester/Plots/histogram_distance_all_vs_cycling_potential.png",
  "data/Manchester/Plots/histogram_distance_cycling_potential_vs_current.png"
))
```

```{r desirefacetcycling, fig.cap="Current and Potential Cycling Demand", out.width="65%", fig.pos="H", fig.align="center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/desire_facet_cycling.png")
```

@@

Figure \@ref(fig:flowsfacetweighted) shows that routing on the weighted
network significantly reduces flow on the trunk and primary roads, but
does not eliminate it completely. This is intentional, as the impedance
on these roads is only slightly higher than remaining road types (See
Table \@ref(table:weightprofiles)). Potential cycling demand is only
routed on these roads if there are no routes through other roads that
offer comparable directness.

@@

@@

The first algorithm focuses on connectivity and directness, but not on
fairly distributing investment. The latter is not a requirement for
increasing cycling uptake, but it is fundamental for spatial equity, as
explained in Section \@ref(ethical-underpinnings-and-proposed-approach). This algorithm incorporates
the ideal of fair distribution by using community detection to partition
the road network.

The algorithm uses the following logic to ensure fair distribution
between communities:

@@

library(tidyverse)
library(sf)

@@

knitr::kable(table,
             caption = "Total Person-Km in Different Communities (Manchester)")

@@

@@

A core part of the methodology is determining where there is a high
potential for cycling, and using this as a basis for recommendations on
road space reallocation in order to unlock potential cycling demand. The
framework provides a routing engine to study the effects of limiting
access on different road types, finding that reducing cyclist flow on
roads with high through-traffic results in acceptable increases in
commuting distances. As a result, a hierarchy of road preference needs
to be defined to route potential cycling demand. The framework provides
different algorithms to determine where investments in cycling
infrastructure should be prioritized. These algorithms, based on
connectivity and egalitarian principles of resource distribution,
ensured that whatever the level of investment, the resulting cycling
network would improve connectivity of existing cycling infrastructure.

The approach is not without limitations. The level of detail is only as
good as the granularity of the available data (in this case relatively
coarse zones). @iacono2010measuring note that such large travel zones
are not ideal for understanding route choice behaviour of cyclists and
pedestrians. They also give rise to an \`ecological fallacy' whereby
average characteristics are assumed to apply to all residents of the
aggregated geographical area, suggesting a need for applying the methods
to more granular OD data (and for governments and other
data-collecting organisations to make OD data more
readily available).

>>>>

*Optimization* techniques have been used to propose improvements to
cycling networks. @mesbah2012bilevel propose a bi-level formulation to
optimize allocation of cycling lanes to the network without exceeding a
set budget. They account for the effect of cycling lanes on car traffic,
and attempt to maximize utilization of said lanes with minimal impact on
car travel times. Safeguarding against increased car traffic may be counter-productive if the goal is to create a mode-shift, as research has shown that reducing road space for cars leads to less cars on the road, a phenomenon referred to as "traffic evaporation" [@nello2020environmental].
Car usage is not considered by @mauttone2017bicycle,
who develop an optimization framework that aims to minimize the total
user cost of cycling on the network. The aggregate flow^[*flow* is
used throughout this research to refer to the cycling demand when it is
routed onto the road network. The flow on any road segment is the
cumulative demand on it, resulting from cyclists commuting between
various OD pairs] on links^[*link*
refers to a road segment throughout this research] is obtained by using
shortest paths to route existing cycling demand onto the road network,
and the solution is a proposed set of links where cycling infrastructure
should be added in order to minimize the overall travel cost of cyclists
across the network. The cost of traversing a link is given as a function
of its length and whether or not it has cycling infrastructure. The
problem has also been solved by attempting to find the minimum cost of
improving roadway links to meet a desired level of service (LOS)
[@duthie2014optimization]. In this formulation, all origin-destination
(OD) pairs need to be connected by roads that meet the desired LOS. A
limitation of these problem formulations is that they do not explicitly solve for
continuity, which is dealt with using either a constraint specifying
that each link with a bike lane should be connected to at least one
destination [@mesbah2012bilevel], a constraint on maximum deviation
from shortest paths [@duthie2014optimization], or a discontinuity
penalty to prioritize connected road segments [@mauttone2017bicycle].

@@

All of these network-level methodologies are underpinned by ethical
principles, even though these principles are not explicitly acknowledged
by the authors. This is important since different ethical principles
constitute different problem formulations and targets. Broadly speaking,
transport appraisal can be based on either utilitarian or egalitarian
principles. The former seeks to maximize the overall benefit, while the
latter is concerned with a fair distribution of benefits
[@jafino2020transport]. 
<!-- @nahmias2017integrating criticize the -->
<!-- utilitarian approach that has been historically popular in the -->
<!-- evaluation of transport investments, explaining how the maximization of -->
<!-- overall benefit fails to account for the distribution of that benefit -->
<!-- among communities or individuals. @lucas2016method explain how -->
<!-- transport studies have traditionally looked at the bigger picture -->
<!-- without studying the distribution of investments on the different parts -->
<!-- of the study area, and go on to propose an egalitarian approach that -->
<!-- ensures the disaggregation of transport policy benefits across the study -->
<!-- area.  -->
The utilitarian approach, historically popular in transport planning, has been criticised for focusing on the bigger picture and failing to account for the distribution of investments on the different communities of the study area [@lucas2016method; @nahmias2017integrating].
@pereira2017distributive emphasize the need for a more
egalitarian approach to transport planning. They highlight accessibility
as a cornerstone of distributive justice, and contend that policies
should aim to distribute investments in a way that minimizes spatial
variations in accessibility.

@@

```{r potdemhistograms, fig.cap="Distribution of Potential Cycling Demand", out.width="30%", fig.show='hold', echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/histogram_distance_all_vs_cycling.png",
  "data/Manchester/Plots/histogram_distance_all_vs_cycling_potential.png",
  "data/Manchester/Plots/histogram_distance_cycling_potential_vs_current.png"
))
```

```{r desirefacetcycling, fig.cap="Current and Potential Cycling Demand", out.width="60%", fig.pos="H", fig.align="center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/desire_facet_cycling.png")
```

@@

Figure \@ref(fig:flowsfacetweighted) shows that routing on the weighted
network significantly reduces flow on the trunk and primary roads, but
does not eliminate it completely. This is intentional, as the impedance
on these roads is only slightly higher than the remaining road types (See
Table \@ref(table:weightprofiles)). Potential cycling demand is only
routed on these roads if there are no routes through other roads that
offer comparable directness.

@@

The algorithm logic is as follows:

@@

The first algorithm focuses on connectivity and directness, but not on
fairly distributing investment. The latter is not a requirement for
increasing cycling uptake, but it is fundamental for spatial equity, as
explained in Section \@ref(ethical-underpinnings-and-proposed-approach). This algorithm incorporates
the ideal of fair distribution by ensuring that investment is distributed between the defined communities. 
This is done using the following logic:

@@

library(tidyverse)
library(sf)
library(kableExtra)

@@

knitr::kable(table,
             caption = "Total Person-Km in Different Communities (Manchester)") %>%
  kable_styling(latex_options = "hold_position")

@@

Overall, the improved distribution of investment shown by Algorithm 2 does not seem to come at the cost 
of poorer connectivity or less city-wide gains when compared to Algorithm 1. 
The results therefore advocate for the incorporation of egalitarian principles in bicycle network planning.

@@

A core part of the methodology is determining where there is a high
potential for cycling, and using this as a basis for recommendations on
road space reallocation in order to unlock potential cycling demand. The
framework provides a routing engine to study the effects of limiting
access on different road types, finding that reducing cyclist flow on
roads with high through-traffic results in acceptable increases in
commuting distances. As a result, a hierarchy of road preference needs
to be defined to route potential cycling demand. The framework provides
an algorithm to determine where investments in cycling
infrastructure should be prioritized. This algorithm, based on
connectivity and egalitarian principles of resource distribution,
ensured that whatever the level of investment, the resulting cycling
network would improve connectivity of existing cycling infrastructure.
The algorithm was also shown to provide similar city-wide results as well better distribution of investment when compared to one based on utilitarian principles.

The approach is not without limitations. The level of detail is only as
good as the granularity of the available data (in this case relatively
coarse zones). @iacono2010measuring note that such large travel zones
are not ideal for understanding route choice behaviour of cyclists and
pedestrians. They also give rise to an \`ecological fallacy' whereby
average characteristics are assumed to apply to all residents of the
aggregated geographical area, suggesting a need for applying the methods
to more granular OD data (and for governments and other
data-collecting organisations to make OD data more
readily available).
The potential demand calculation is also based on cycling in the traditional sense, and does not consider the 
effect of micro-mobility on reducing topology-related impedance to cycling. Given that we are proposing a methodological framework which can accommodate any cycling uptake functions, this is an acceptable limitation.
